Parsing Postfix log files

### 2008-04-01 doc/logparser.tex
Corrections from Carl for Sections 1 and 2.

% When creating a \PDF{} make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
\ifpdf{}
    \usepackage[pdftex]{hyperref}
\else{}
    \usepackage[dvips]{hyperref}
\fi{}
\hypersetup{
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
%    breaklinks  = true,
}
% This is necessary for URLs in the bibliography; I dunno why the url
% package doesn't work; don't include it when generating \PDF{} output.
\ifpdf{}
\else{}
    \usepackage{breakurl}
\fi{}
\usepackage{lastpage}

@@

@@

Most mail server administrators will have performed some basic processing
of the logs produced by their mail server at one time or another, whether
it was to debug a problem, explain to a user why their mail is being
rejected, or check whether new anti-spam measures are working.  The more
adventurous will have generated statistics to show how many hits each of
their anti-spam measures has gotten in the last week, and possibly even
generated some graphs to clearly illustrate the point to management or
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this type of processing is the
unstructured nature of Postfix logs, where each log line was added on an ad
hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from \url{http://www.postfix.org/}} Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every \RBL{}
returns a different explanatory message; policy servers may log different
messages depending on the characteristics of the connection; there are many
ways in which the log lines may differ between servers, even within the
same organisation --- servers may be configured differently, or running
different version of Postfix.  This paper documents the difficult process
of parsing Postfix logs, presenting a program which parses logs and places
the resulting data into a database for later use.  The gathered data can
then be used to optimise current anti-spam measures, provide a baseline to
test new anti-spam measures against, or to produce statistics showing how
effective those measures are.  There are numerous other uses for such data:
improving server performance by identifying troublesome destinations and
reconfiguring appropriately; identifying regular high volume uses (e.g.\
customer newsletters) and restricting those uses to off-peak times;
detecting virus outbreaks which propagate via email; as a base for billing
customers on a shared server.  Preserving the raw data enables users to
develop a multitude of uses far beyond those conceived of by the author.

@@

@@

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are beneficial in the fight against spam.  The
program parses Postfix logs and populates a database with the data gleaned
from those logs, providing a consistent and simple view of the logs which
future tools can utilise.  The gathered data can then be used to optimise
current anti-spam measures, provide a baseline to test new anti-spam
measures against, or to produce statistics showing how effective those
measures are.\footnote{\sectionref{introduction} lists some more example
uses.}

@@

The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and viruses.\footnote{This violates
the principle of \textit{Be liberal in what you accept, and conservative in
what you send\/} from \RFC{}~760~\cite{rfc760}, but unfortunately that
principle was written in a friendlier time.  Given the deluge of spam that
mail servers are subjected to daily, a more appropriate maxim could be:
\textit{Require strict adherence to \RFC{}~2821; implement the strongest
restrictions you can; relax the restrictions and adherence only when
legitimate mail is impeded.\/}  It's not as friendly, nor as catchy, but it
more accurately reflects the current situation.} A typical \SMTP{}
conversation resembles the following (the lines starting with a three digit
number are sent by the server, all other lines are sent by the client):

@@

This client is so poorly written that not only does it present the sender
and recipient addresses improperly, it ignores the error messages returned
by the server and carries on regardless.  There are many spam and virus
sending programs which are this deficient --- unfortunately others
(particularly newer programs) were written by competent programmers, or
utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
hosts, Microsoft Outlook on Windows hosts).

@@

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
into various component programs to perform the tasks required of an
\MTA{}\@: receive mail, send mail, local delivery of mail, etc. --- full
details can be found at~\cite{postfix-overview}.  Each log line contains
the name of the Postfix program which produced it, and this information is
used when determining which rules should be used to parse each log line
(see \sectionref{rule characteristics} for details.)  This design has
positive security implications also: those components which interact with
other hosts are not privileged, and those components which are privileged
do not interact with other hosts.

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an \SMTP{} command, one list for each
stage of the \SMTP{} conversation: client connection, HELO command, MAIL
FROM command, RCPT TO command (possibly multiple times), DATA command, and
end of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a benefit of this delay is that Postfix has
more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter in Postfix's configuration controls
this behaviour.}  Postfix uses simple lookup tables as the deciding factor
for some restrictions, e.g.\newline
\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\newline The line above can be broken down as follows:

@@

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible, for full details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue processing the remainder of the restriction list, allowing more
specific cases to be used as exceptions to more general cases.  When the
result is the name of another restriction Postfix will evaluate the new
restriction, allowing restrictions to be chosen based on the client \IP{}
address, HELO hostname, sender address, recipient address, etc.  The
administrator can define new restrictions as a list of existing
restrictions, allowing arbitrarily long and complex sequences of lookups,
restrictions and exceptions.

@@

    \item Rate limiting or throttling on a per-sender, per-client or
        per-recipient basis as performed by Policyd~\cite{policyd}.

@@

    \item The Postfix queue has not been tampered with, causing unexplained
        appearance or disappearance of mail.

@@

    \item The source code cannot accurately communicate the variety of
        orderings in which log lines are written to the log file, as
        process scheduling interferes with logging and other processing.

@@

    \item [rules] These match individual log lines and determine which
        actions will be executed.  Rules provide an easily extensible
        method of associating log lines and actions.  Rules are described
        in detail in \sectionref{rules}.

    \item [actions] Are invoked to deal with a log line once it has been
        identified by the rules: modifying data structures, handing
        complications, and saving data to the database.  The actions
        perform the work of reconstructing the journey a mail takes through
        Postfix.  Full details of actions can be found in
        \sectionref{parsing-algorithm}, particularly
        \sectionref{actions-in-detail} and \sectionref{adding new actions}.

    \item [framework] Is responsible for loading rules, managing data
        structures so actions can retrieve them, finding the rule which
        matches each log line, invoking the correct action, etc.  The
        framework provides the structure which actions and rules plug into.
        The framework is not described in detail in this document because
        it is not particularly interesting, but there is some coverage in
        \sectionref{framework}.

@@

Separating the rules from the actions and framework makes it possible to
parse new log lines without modifying the core parsing algorithm.  Although
this may seem like a trivial point, is it substantially more difficult to
understand a program's entire parsing algorithm, identify the correct
location to change and make the appropriate changes, versus adding a new
rule with the action to invoke, a \regex{} to match the log lines, and the
specification of the data to extract.  Bear in mind that the changes must
be made without adversely affecting existing parsing, particularly as there
may be edge cases which are not immediately obvious.\footnote{See
\sectionref{yet-more-aborted-delivery-attempts} for a complication which
occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
changes to the parser's code also complicates upgrades, as the changes
must be preserved during the upgrade, and may clash with changes made by
the developer.\footnote{See \sectionref{adding new actions} for how to add
new actions}.  \parsername{} allows the user to add new rules to the
database without changing the parsing algorithm, unless the new log lines
to be parsed require functionality not already provided by the algorithm.
If the new log lines do require new functionality, new actions can be added
to the parser without modifying existing actions or other parts of the
algorithm; only in the rare case that the new actions require support from
other sections of the code will more extensive changes be required.

@@

Berkeley DB is not an \SQL{} database, supporting only \textbf{(key,
value)} pairs; in the main table the key is an integer referred to by
secondary tables and the value used is a \CSV{} line containing all the
data for that row.  The secondary by-sender, by-recipient, by-date and
by-\IP{} tables use the sender/recipient/date/\IP{} as the key, and a
\CSV{} list of integers referring to the main table is the value.  In
effect this re-implements \SQL{} foreign keys, but without the
functionality offered by even the most basic of \SQL{} databases (joins,
ordering, searches, etc.).  It also requires custom code to search on some
combination of the above, though the authors of \LMA{} did provide some
queries: IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB
appears to be the least useful of the three output formats

@@

Notes about the command above: there is an unnecessary line containing only
a single space output after each matching line, for compatibility with the
output of DAILY-REJECT\@.  Some records have an \SMTP{} code of 0, though
this is not a valid \SMTP{} code, so presumably this is due to \LMA{} not
parsing a line properly; \LMA{} considers this code to be a rejection, so
the command above must too.  In a tiny proportion of records the \SMTP{}
code is \textit{deferred\/} (again, not a valid code, and most likely due
to mis-parsing), and is treated as a rejection by \LMA{}, so the equivalent
\texttt{awk} command must do so too.

@@

    \item A description of the parser's novel three section design.

@@

The database is an integral part of the parser: it stores the rules and the
data gleaned by applying those rules to Postfix log files.  Understanding
the database schema is important in understanding the actions of the
parser, and essential to developing further applications which utilise the
data gathered.

>>>>

% When creating a \PDF{} make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
% hyperfootnotes need to be disabled to avoid breaking footmisc, but they
% still seem to work, somehow.
\ifpdf{}
    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
\else{}
    \usepackage[dvips,hyperfootnotes=false]{hyperref}
\fi{}
\hypersetup{
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
%    breaklinks  = true,
}
% This is necessary for URLs in the bibliography; I dunno why the url
% package doesn't work; don't include it when generating \PDF{} output.
\ifpdf{}
\else{}
    \usepackage{breakurl}
\fi{}
\usepackage{lastpage}

% Extra footnote functionality, including references to earlier footnotes.
\usepackage{footmisc}

@@

XXX PICK A NAME AND INTRODUCE IT\@.

@@

Most mail server administrators will have performed some basic processing
of the logs produced by their mail server at one time or another, whether
it was to debug a problem, explain to a user why their mail is being
rejected, or check whether new anti-spam measures are working.  The more
adventurous will have generated statistics to show how many hits each of
their anti-spam measures has gotten in the last week, and possibly even
generated some graphs to clearly illustrate the point to management or
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this type of processing is the
unstructured nature of Postfix logs, where each log line was added on an ad
hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from \url{http://www.postfix.org/}} Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every
\RBL{}\footnote{This document is supplied with a glossary, see
appendix~\ref{Glossary}.} returns a different explanatory message; policy
servers may log different messages depending on the characteristics of the
connection; there are many ways in which the log lines may differ between
servers, even within the same organisation --- servers may be configured
differently, or running different version of Postfix.  This paper documents
the difficult process of parsing Postfix logs, presenting a program which
parses logs and places the resulting data into a database for later use.
The gathered data can then be used to optimise current anti-spam measures,
provide a baseline to test new anti-spam measures against, or to produce
statistics showing how effective those measures are.  There are numerous
other uses for such data: improving server performance by identifying
troublesome destinations and reconfiguring appropriately; identifying
regular high volume uses (e.g.\ customer newsletters) and restricting those
uses to off-peak times; detecting virus outbreaks which propagate via
email; as a base for billing customers on a shared server.  Preserving the
raw data enables users to develop a multitude of uses far beyond those
conceived of by the author.

@@

XXX HOW MANY MAILS PER DAY\@?

@@

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are beneficial in the fight against spam.  The
program parses Postfix logs and populates a database with the data gleaned
from those logs, providing a consistent and simple view of the logs which
future tools can utilise.  The gathered data can then be used to optimise
current anti-spam measures, provide a baseline to test new anti-spam
measures against, or to produce statistics showing how effective those
measures are.

@@

The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and
viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
adhered to the principle of \textit{Be liberal in what you accept, and
conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
unfortunately that principle was written in a friendlier time.  Given the
deluge of spam that mail servers are subjected to daily, a more appropriate
maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
the strongest restrictions you can; relax the restrictions and adherence
only when legitimate mail is impeded.\/}  It's not as friendly, nor as
catchy, but it more accurately reflects the current situation.} A typical
\SMTP{} conversation resembles the following (the lines starting with a
three digit number are sent by the server, all other lines are sent by the
client):

@@

This client is so poorly written that not only does it present the sender
and recipient addresses improperly, it ignores the error messages returned
by the server and carries on regardless.  There are many spam and virus
sending programs which are this deficient --- unfortunately others
(particularly newer programs) were written by competent programmers, or
utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
would have done its best to deal with deficient clients, with the intention
of accepting as much mail destined for its users as
possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
command, or accepting sender or recipient addresses which were not enclosed
in \texttt{<>}.

@@

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
principle was also followed when designing the parser (\sectionref{parser
design}), notably having separate actions rather than one huge action.} and
is separated into various component programs to perform the tasks required
of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
full details can be found at~\cite{postfix-overview}.  Each log line
contains the name of the Postfix program which produced it, and this
information is used when determining which rules should be used to parse
each log line (see \sectionref{rule characteristics} for details.)  This
design has positive security implications also: those components which
interact with other hosts are not privileged, and those components which
are privileged do not interact with other hosts.

@@

Postfix uses one restriction list (containing zero or more restrictions)
for each stage of the \SMTP{} conversation: client connection, HELO
command, MAIL FROM command, RCPT TO commands, DATA command, and end of
data.  The appropriate restriction list is evaluated for each stage
(evaluation will be explained shortly), though by default the restriction
lists for client connection, HELO and MAIL FROM commands will not be
evaluated until the first RCPT TO command is received, because some clients
don't deal properly with rejections before RCPT TO\@; a benefit of this
delay is that Postfix has more information available when logging the
rejection.  Postfix uses simple lookup tables as the deciding factor when
evaluating some restrictions, e.g.\newline
\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\newline The line above can be broken down as follows:

@@

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible, for full details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue processing the remainder of the restriction list, allowing more
specific cases to be used as exceptions to more general cases.  When the
result is the name of another restriction Postfix will evaluate the new
restriction, allowing restrictions to be chosen based on the client \IP{}
address, HELO hostname, sender address, recipient address, etc.  The
administrator can define new restrictions as a list of existing
restrictions, allowing arbitrarily long and complex sequences of lookups,
restrictions and exceptions.  Postfix tries to protect the administrator in
as far as is reasonable, e.g.\ the restriction
\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
because the parameter it checks (the hostname given in the HELO command) is
under the control of the remote client.  Despite this, it is possible for
the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
all users --- the administrator must remember that \textit{With great
power comes great responsibility}.

@@

    \item Rate limiting on a per-sender, per-client or per-recipient basis
        as performed by Policyd~\cite{policyd}.

@@

    \item The Postfix queue has not been tampered with, causing unexplained
        appearance or disappearance of mail.  This may happen if the
        administrator deletes mail from the queue without using 
        \daemon{postsuper}, or if there is filesystem corruption.

@@

    \item The source code cannot accurately communicate the variety of
        orderings in which log lines are written to the log file, as
        process scheduling independently interferes with logging and other
        processing.

@@

    \item [rules] Rules match individual log lines and determine which
        actions will be executed.  Rules provide an easily extensible
        method of associating log lines and actions.  Rules are described
        in detail in \sectionref{rules}.

    \item [actions] Actions invoked to deal with a log line once it has been
        identified by the rules: modifying data structures, handing
        complications, and saving data to the database.  The actions
        perform the work of reconstructing the journey a mail takes through
        Postfix.  Full details of actions can be found in
        \sectionref{parsing-algorithm}, particularly
        \sectionref{actions-in-detail} and \sectionref{adding new actions}.

    \item [framework] The framework is responsible for loading rules, managing data
        structures so actions can retrieve them, finding the rule which
        matches each log line, invoking the correct action, etc.  The
        framework provides the structure which actions and rules plug into.
        The framework is not described in detail in this document because
        it is not particularly interesting, but there is some coverage in
        \sectionref{framework}.

@@

Separating the rules from the actions and framework makes it possible to
parse new log lines without modifying the core parsing algorithm.  Although
this may seem like a trivial point, is it substantially more difficult to
understand a program's entire parsing algorithm, identify the correct
location to change and make the appropriate changes, versus adding a new
rule with the action to invoke, a \regex{} to match the log lines, and the
specification of the data to extract.  Bear in mind that the changes must
be made without adversely affecting existing parsing, particularly as there
may be edge cases which are not immediately obvious.\footnote{See
\sectionref{yet-more-aborted-delivery-attempts} for a complication which
occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
changes to the parser's code also complicates upgrades, as the changes
must be preserved during the upgrade, and may clash with changes made by
the developer.\footnote{See \sectionref{adding new actions} for how to add
new actions}  \parsername{} allows the user to add new rules to the
database without changing the parsing algorithm, unless the new log lines
to be parsed require functionality not already provided by the algorithm.
If the new log lines do require new functionality, new actions can be added
to the parser without modifying existing actions or other parts of the
algorithm; only in the rare case that the new actions require support from
other sections of the code will more extensive changes be required.

@@

Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
\SQL{} databases which store arbitrary tuples; in the main table the key is
an integer referred to by secondary tables, and the value used is a \CSV{}
line containing all of the data for that row.  The secondary by-sender,
by-recipient, by-date and by-\IP{} tables use the
sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
referring to the main table is the value.  In effect this re-implements
\SQL{} foreign keys, but without the functionality offered by even the most
basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
requires custom code to search on some combination of the above, though the
authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
of the three output formats

@@

Notes about the command above:

    \item It outputs an unnecessary line containing only a single space
        after each matching line, for compatibility with the output of
        DAILY-REJECT\@.

    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
        line properly.  \LMA{} considers this code to be a rejection, so
        the command above must, too.

    \item In a tiny proportion of records the \SMTP{} code is
        \textit{deferred\/} (again, not a valid code, and most likely due
        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
        so the equivalent \texttt{awk} command must do so, too.

@@

    \item A description of a new parser's novel three section design.

@@

The database is an integral part of the parser presented here: it stores
the rules and the data gleaned by applying those rules to Postfix log
files.  Understanding the database schema is important in understanding the
actions of the parser, and essential to developing further applications
which utilise the data gathered.
